### 2008-12-17 talk/ecoop2009/clibackend.tex
management of external links almost finished

Since in .NET methods are the basic units of compilation, a possible
solution consists in creating a new method 
any time a new case has to be added to a flexswitch.
\dacom{comment for Antonio: I am not sure this is the best solution. This cannot work for Java where classes are the basic
  units. Closures will be available only with Java Dolphin and I do
  not know how much efficient will be}
In this way, whereas flow graphs without flexswitches are translated
to a single method, the translation of flow graphs which can dynamically grow because of
flexswitches will be scattered over several methods.
Summarizing, the backend behaves in the following way:
\begin{itemize}
\item Each flow graph is translated in a collection of methods which
  can grow dynamically. \dacom{I propose primary/secondary instead of
    the overloaded terms main/child} Each collection must contain at least one
  method, called \emph{primary}, which is the first to be created.
  All other methods, called \emph{secondary}, are added dynamically 
  whenever a new case is added to a flexswitch.

\item Each either primary or secondary method corresponds to the
  translation of some of the blocks of a single flow graph. Each
  method has an initial block whose input variables are the
  parameters of the method; the input variables of all other blocks
  are local variables of the method.
\end{itemize}

@@

A link is called \emph{internal} if it connects two blocks implemented in the same method,
 \emph{external} otherwise.

Following an internal link is not difficult in CLI bytecode: a jump to
the corresponding code fragment in the same method is emitted 
to execute the new block, whereas the appropriate local variables are
used for passing arguments.
Also following an external link whose target is the initial block of a
method is not difficult: the corresponding method has to be invoked
with the appropriate arguments.

@@

To solve this problem a special block, called \emph{dispatcher}, is
added to every method; whenever a method is invoked, its dispatcher is
executed first\footnote{Recall that the dispatcher is a special block
and must not be confused with the initial block of a method.} to
determine which block has to be executed.
This is done by passing to the method a 32 bits number, called 
\emph{block id}, which uniquely identifies the next block of the graph to be executed.
The high word of a block id is the id of the method to which the block
belongs, whereas the low word is a progressive number identifying
each block implemented by the method.

@@

\commentout{
Each method in a graph is assigned an unique 16 bit method id; each block in a method is assigned a progressive 16 bit block number. From this two numbers, we can compute the block id as an unsigned integer, by storing the method id in the first 16 bits and the block number in the second 16 bits. By construction, the block id is guaranteed to be unique in the graph.

The following picture shows a graph composed of three methods; the id of each method is shown in red, while the block ids are shown in red (for the method id part) and black (for the block number part). The graph contains three external links; in particular, note the link between blocks 0x00020001 and 0x00010001 which connects two block that resides in different methods.

Every method contains a special dispatch block, (not shown in the picture above) whose goal is to jump to the specified block number inside the method itself. The first argument of a secondary method is always a block id; when the method starts, it immediately jumps to the dispatch block, and thus to the desired block.

For example, suppose to have a method which contains 3 blocks numbered 0, 1, 2; here is how its dispatch blocks looks like; for simplicity it is shown as C# code, but it is actually generated as IL bytecode:

// dispatch block
int methodid = (blockid & 0xFFFF0000) >> 16); // take the first 16 bits
int blocknum = blockid && 0x0000FFFF;         // take the second 16 bits

if (methodid != MY_METHOD_ID) {
// jump_to_unknown block
...
}

switch(blocknum) {
case 0:
goto block0;
case 1:
goto block1;
case 2:
goto block2;
default:
throw new Exception("Invalid block id");
}

Whenever we want to jump to a external block, it is enough to store the block id in the appropriate variable and jump to the dispatch block. If the block resides in a different method, the jump_to_unknown block is entered; this special block is implemented differently by the main method and the secondary methods, as we will see soon.

Each time a new method is added to the graph, we build a delegate for it, and store it in a special array called method_map; since we assign the method id sequentially starting from 0, we are sure that to fetch the method whose id is n we can simply load the n-th element of the array.

The jump_to_unknown block of the main method uses this array to select the right method, and calls it (FlexSwitchCase is the type of delegates for all secondary methods):

// jump_to_unknown block of the main method
FlexSwitchCase meth = method_map[methodid];
blockid = meth(blockid, ...); // execute the method
goto dispatch_block;

Each secondary method returns a block id specifying the next block to jump to; after its execution, we assign the return value to the blockid variable, and jump again to the dispatch block, which will jump again to the appropriate block.

Keeping this in mind, it is straightforward to implement the jump_to_unknown block of secondary methods: it is enough to return the target block id to the caller, and let its dispatch loop do the right thing. If the caller is also a secondary method, it will return it again, until we reach the dispatch loop of the main method, which will finally do the jump. In theory, we could implement things differently and jumping directly from a secondary method to another one, but in that case the call stack could grows indefinitely in case of a tight loop between two blocks residing in different methods.

To implement the dispatch block we can exploit the switch opcode of the CLI; if the .NET JIT is smart enough, it can render it using an indirect jump; overall, jumping to a external block consists of an indirect function call (by invoking the delegate) plus an indirect jump (by executing the switch opcode); even if this is more costly than a simple direct jump, we will see in the next section that this not the main source of overhead when following a external link.

@@

% LocalWords:  flexswitches backend flexswitch

>>>>

Since in .NET methods are the basic units of compilation, a possible
solution consists in creating a new method 
any time a new case has to be added to a flexswitch.
\dacom{comment for Antonio: I am not sure this is the best solution. This cannot work for Java where classes are the basic
  units. Closures will be available only with Java Dolphin and I do
  not know how much efficient will be}
In this way, whereas flow graphs without flexswitches are translated
to a single method, the translation of flow graphs which can dynamically grow because of
flexswitches will be scattered over several methods.
Summarizing, the backend behaves in the following way:
\begin{itemize}
\item Each flow graph is translated in a collection of methods which
  can grow dynamically. \dacom{I propose primary/secondary instead of
    the overloaded terms main/child} Each collection contains at least one
  method, called \emph{primary}, which is the first to be created.
  All other methods, called \emph{secondary}, are added dynamically 
  whenever a new case is added to a flexswitch.

\item Each either primary or secondary method implements a certain
  number of blocks, all belonging to the same flow graph. Among these blocks
  there always exists an initial block whose input variables are
  parameters of the method; the input variables of all other blocks
  are local variables of the method.
\end{itemize}

@@

A link is called \emph{internal} if it connects two blocks implemented
by the same method,
 \emph{external} otherwise.

Following an internal link is not difficult in IL bytecode: a jump to
the corresponding code fragment in the same method is emitted 
to execute the new block, whereas the appropriate local variables are
used for passing arguments.
Also following an external link whose target is the initial block of a
method is not difficult: the corresponding method has to be invoked
with the appropriate arguments.

@@

To solve this problem every method contains a special code, called
\emph{dispatcher}; whenever a method is invoked, its dispatcher is
executed first\footnote{The dispatcher should not be
confused with the initial block of a method.} to
determine which block has to be executed.
This is done by passing to the method a 32 bits number, called 
\emph{block id}, which uniquely identifies the next block of the graph to be executed.
The high word of a block id is the id of the method to which the block
belongs, whereas the low word is a progressive number univocally identifying
each block implemented by the method.

@@

For instance, the code\footnote{For simplicity we write C\# code instead of
the actual IL bytecode.} generated for the dispatcher of method \texttt{0x0002}
is similar to the following fragment: 
\begin{small}
\begin{lstlisting}[language={[Sharp]C}]
// dispatch block
int methodid = (blockid && 0xFFFF0000) >> 16; 
int blocknum = blockid && 0x0000FFFF;         
if (methodid != MY_METHOD_ID) {
  // jump_to_ext 
  ...
}
switch(blocknum) {
  case 0: goto block0;
  case 1: goto block1;
  default: throw new Exception("Invalid block id");
}
\end{lstlisting}
\end{small}
If the next block to be executed is implemented in the same method
({\small\lstinline{methodid == MY_METHOD_ID}}), then the appropriate
jump to the corresponding code is executed. Otherwise, the \lstinline{jump_to_ext}
part of the dispatcher has to be executed.
The code that actually jumps to an external block is contained in
the dispatcher of the primary method, whereas the
\lstinline{jump_to_ext} code of dispatchers of secondary methods
simply delegates the dispatcher of the primary method of the same
graph (see later).

The primary method is responsible for the bookkeeping of the secondary
methods which are added to the same graph dynamically. This can be 
simply implemented with an array mapping method id of secondary methods
to the corresponding delegate. Therefore, the primary methods contain
the following \lstinline{jump_to_ext} code (where
\lstinline{FlexSwitchCase} is the type of delegates for secondary methods):
\begin{small}
\begin{lstlisting}[language={[Sharp]C}] 
// jump_to_ext
FlexSwitchCase meth = method_map[methodid];
blockid = meth(blockid, ...); // execute the method
goto dispatch_block;
\end{lstlisting}
\end{small}
Each secondary method returns the block id of the next block to be
executed; therefore, after the secondary method has returned, the
dispatcher of the primary method will be executed again to jump
to the correct next block.

To avoid mutual recursion and an undesired growth of the stack,
the \lstinline{jump_to_ext} code in dispatchers of secondary methods
just returns the block id of the next block; since the primary method
is always the first method of the graph which is called, the correct
jump will be eventually executed by the dispatcher of the primary method.

\commentout{
To implement the dispatch block we can exploit the switch opcode of the CLI; if the .NET JIT is smart enough, it can render it using an indirect jump; overall, jumping to a external block consists of an indirect function call (by invoking the delegate) plus an indirect jump (by executing the switch opcode); even if this is more costly than a simple direct jump, we will see in the next section that this not the main source of overhead when following a external link.

@@

% LocalWords:  flexswitches backend flexswitch methodid blockid xFFFF blocknum
% LocalWords:  FFFF goto FlexSwitchCase meth
